Ecosystem size and complexity as extrinsic drivers of food chain length in branching ecosystems

### 2023-02-24 rmd/manuscript.Rmd
edit: manuscript, method

# Sensitivity data
load(file = here::here("output/result_main.RData"))

n_rep <- max(sim_main_result$n_rep)

n_scenario <- sim_main_result %>% 
  filter(sd_disturb_source > sd_disturb_lon) %>% 
  summarize(n = n_distinct(param_set)) %>% 
  pull(n)

n_sim <- sim_main_result %>% 
  mutate(n_sim = n_timestep + n_burnin + n_warmup) %>% 
  pull() %>% 
  unique()
```

@@

No data were used or produced in this research. R functions for simulations are provided as the R package mcbrnet version 1.2.3 (available at <https://github.com/aterui/mcbrnet>). Codes for simulations, statistical analysis, figures, and tables are available at <https://github.com/aterui/public-proj_fcl-river-theory>.

@@

## **Tri-trophic dynamic simulation model**

@@

***Internal dynamics.*** Each time step starts with internal dynamics. The general formula is given as follows:

@@

where

@@

The parameter $e_{ij}$ is the energetic conversion efficiency from $i$ to $j$. The function $g(\cdot)$ defines the density-dependent recruitment for the basal species, and the function $f_k(\cdot)$ dictates the proportion of survived prey individuals after predation. We used the Beverton-Holt function to describe the density-dependent growth of the basal species:

@@

The parameter $a_{ij}$ defines the encounter rate between $j$ and $i$, $h_{ij}$ the handling time, and $\phi$ the switching function for the intraguild predator. We assumed that the degree of omnivory is dependent on the relative abundance of $B''_{x,t}$ and $C'_{x,t}$:

$$
\phi = \frac{s(B''_x(x,t) - C'_{x,t})}{B''_x(x,t) + C'_{x,t}}
$$

@@

***Dispersal.*** We simulated distance-dependent dispersal along with a network. The number of emigrants from patch $x$ at time $t$, $E_{i,x}(t)$, is estimated as the product of dispersal probability $p_{d}$ and species abundance after internal dynamics $\bar{A}_{i,x}(t)$. The dispersal probability was assumed to be constant for all three species. We calculated the relative immigration potential $\xi_{i,x}(t)$ considering network structure and species dispersal capability as follows:

$$
\begin{aligned}
\xi_{i,x}(t) &= \frac{\sum_{y, y \ne x}E_{i,y}(t)\exp(-\theta d_{xy})}{\sum_{x} \sum_{y, y \ne x}\text{E}_{i,y}(t)\exp(-\theta d_{xy})}
&&\text{(16)}
\end{aligned}
$$

where $d_{xy}$ is the network distance between patch $x$ and $y$ (shortest path), and $\theta$ indicates the inverse of mean dispersal distance (i.e., larger numbers mean shorter dispersal distance). The immigration potential at patch $x$, $\sum_{y, y \ne x}E_{i,y}(t)\exp(-\theta d_{xy})$, is scaled by the sum of immigration potential across patches so that $\sum_{x} \xi_{i,x}=1$. The expected number of immigrants into patch $x$ is calculated as $I_{i,x}(t)=\xi_{i,x}(t) \sum_{x}E_{i,x}(t)$.

***Food chain length.*** At each patch and time step, we determined FCL as the highest trophic position ($TP_{x}(t)$) of the existing species. When the predator exists, the trophic position is determined as follows:

$$
\begin{aligned}
TP_{P,x}(t) &= \delta_{x}(t) + 2(1 - \delta_{x}(t)) + 1
&&\text{(17)}
\end{aligned}
$$

This definition accounts for the influence of omnivory.

@@

Prior to the main simulation, we performed a one-patch simulation to explore the predation parameter space ($a_{ij}$ and $h_{ij}$) that allows stable coexistence of the three species in the absence of disturbance. In this simulation, we assumed $h_{ij} = h$ to reduce the dimension of parameter space. This analysis revealed that the stable coexistence of the three species is possible with a certain combinations of predation parameters. Within this parameter space, we explored the three scenarios of food web structure: "chain" ($(a_{BC}, a_{BP}, a_{CP}, h) = (X, X, X, X)$), "weak omnivory" ($(a_{BC}, a_{BP}, a_{CP}, h) = (X, X, X)$), and "strong omnivory" ($(a_{BC}, a_{BP}, a_{CP}, h) = (X, X, X)$). We used $s = 1$ except for the chain scenario with $s = 0$. Without the loss of generality, we assumed $e_{ij} = e = 1$.

We controlled disturbance level with disturbance probability $p_{m}$ (low disturbance, `r min(sim_main_result$p_disturb)`; high disturbance, `r max(sim_main_result$p_disturb)`). We used the identical value of mean disturbance intensity at headwaters ($\mu_{m}=$ `r max(sim_main_result$mean_disturb_source)`), spatial variation in disturbance intensity at headwaters ($\sigma_{h}=$ `r max(sim_main_result$sd_disturb_source)`), and local disturbance variation ($\sigma_{l}=$ `r min(sim_main_result$sd_disturb_lon)`) regardless of disturbance scenario. This setup ($\sigma_h > \sigma_l$) produces high spatial variation among tributaries with strong within-tributary downstream correlation, a typical pattern in real river networks [@nakamuraDisturbanceRegimesStream2000; @bendaNetworkDynamicsHypothesis2004]. Productivity was varied by setting different values of basal species' growth rate ($r_{b}=$ `r unique(sim_main_result$r_b)`). For dispersal, we examined short- and long-distance dispersal scenarios (short-distance dispersal $\theta=$ `r max(sim_main_result$theta)`; long-distance dispersal $\theta=$ `r min(sim_main_result$theta)`) with constant dispersal probability ($p_{d}=$ `r unique(sim_main_result$p_dispersal)`).

Under each scenario, we ran `r max(sim_main_result$n_rep)` simulation replicates with `r n_sim` timesteps of trophic dynamics (including `r n_sim - unique(sim_main_result$n_timestep)` steps of warmup and burn-in). During the warmup period (`r unique(sim_main_result$n_warmup)` time steps), we introduced propagules into habitat patches every 10 time steps to allow species to establish populations. Propagules are random draws from $\text{Poisson}(0.5)$. Subsequently, we ran a burn-in period with no propagules for `r unique(sim_main_result$n_burnin)` time steps to remove the effects of initial conditions. We obtained the mean FCL across all habitat patches for the last `r unique(sim_main_result$n_timestep)` time steps. Note that simulation replicates are differ in ecosystem size ($N_{p}\sim \text{Unif}(10,150)$) and ecosystem complexity ($P_{b} \sim \text{Unif}(0.01, 0.99)$) so we can estimate their effects on FCL. All analysis was performed in R 4.1.0 [@rcoreteamLanguageEnvironmentStatistical2021]

>>>>

# Sensitivity data
df_sim <- readRDS(here::here("output/sim_main.rds")) %>% 
  filter(r_b %in% c(8, 16))

n_rep <- max(df_sim$n_rep)

n_scenario <- df_sim %>% 
  filter(sd_disturb_source > sd_disturb_lon) %>% 
  summarize(n = n_distinct(param_set)) %>% 
  pull(n)

n_sim <- df_sim %>% 
  mutate(n_sim = n_timestep + n_burnin + n_warmup) %>% 
  pull() %>% 
  unique()
```

@@

No data were used or produced in this research. R functions for simulations are provided as the R package mcbrnet version 1.3.0 (available at <https://github.com/aterui/mcbrnet>). Codes for simulations, statistical analysis, figures, and tables are available at <https://github.com/aterui/public-proj_fcl-river-theory>.

@@

## **Tri-trophic model**

@@

***Internal dynamics.*** Each time step starts with internal dynamics. The general formula is the extension of the Nicholson-Bailey model, which is given as follows:

@@

where $e_{ij}$ is the energetic conversion efficiency from $i$ to $j$. The function $g(\cdot)$ defines the density-dependent recruitment for the basal species, and the function $f_k(\cdot)$ dictates the proportion of survived prey individuals after predation. In the above equations, we re-defined variables as follows to simplify the expression:

@@

The model assumes that ecological events occur in the following order: (1) $B$'s recruitment, (2) $C$'s predation on $B$ followed by its recruitment, and (3) $P$'s predation on $B$ and $C$ followed by its recruitment. We assumed that the intraguild prey $C$ consumes the basal species $B$ first because the species is specialized to prey on $B$.

We used the Beverton-Holt function to describe the density-dependent growth of the basal species:

@@

The parameter $a_{ij}$ defines $j$'s attack rate to $i$, $h_{ij}$ the handling time, and $\phi$ the switching function for the intraguild predator. We assumed that switching is dependent on the relative abundance of $B''_{x,t}$ and $C'_{x,t}$:

$$
\phi = \frac{s(B''_{x,t} - C'_{x,t})}{B''_{x,t} + C'_{x,t}}
$$

@@

***Dispersal.*** We simulated distance-dependent dispersal along with a network. The number of emigrants from patch $x$ at time $t$, $E_{ix,t}$, is estimated as the product of dispersal probability $p_{d}$ and species abundance after internal dynamics $\bar{A}_{ix,t}$. The dispersal probability was assumed to be constant for all three species. We calculated the relative immigration potential $\xi_{ix,t}$ considering network structure and species dispersal capability as follows:

$$
\begin{aligned}
\xi_{ix,t} &= \frac{\sum_{y, y \ne x}E_{iy,t}\exp(-\theta d_{xy})}{\sum_{x} \sum_{y, y \ne x}E_{iy,t}\exp(-\theta d_{xy})}
&&\text{(16)}
\end{aligned}
$$

where $d_{xy}$ is the network distance between patch $x$ and $y$ (shortest path), and $\theta$ indicates the inverse of mean dispersal distance (i.e., larger numbers mean shorter dispersal distance). The immigration potential at patch $x$, $\sum_{y, y \ne x}E_{iy,t}\exp(-\theta d_{xy})$, is scaled by the sum of immigration potential across patches so that $\sum_{x} \xi_{i,x}=1$. The expected number of immigrants into patch $x$ is calculated as $I_{ix,t}=\xi_{ix,t} \sum_{x}E_{ix,t}$.

***Food chain length.*** At each patch and time step, we determined local FCL as the highest trophic position ($\tau_{x,t}$) of the existing species. When the predator exists, the trophic position is determined as follows:

$$
\begin{aligned}
\tau_{x,t} &= \delta_{x,t} + 2(1 - \delta_{x,t}) + 1
&&\text{(17)}
\end{aligned}
$$

where $\delta_{x,t}$ is the derived parameter that represents the degree of omnivory and is defined as:

$$
\delta_{x,t} = \frac{e_{BP}B'_{x,t}f_1(B'_{x,t}, C_{x,t})[1 - f_2(B''_{x,t}, P_{x,t})]}{P^+_{x,t}}
$$

Thus, our definition accounts for the influence of omnivory. We defined the global FCL $\bar{\tau}$ across patches $N_p$ and time $T$ as $\bar{\tau} = \frac{\sum_x \sum_t \tau_{x,t}}{N_p T}$ (i.e., average over space and time).

@@

Prior to the main simulation, we performed a one-patch simulation to explore the predation parameter space ($a_{ij}$ and $h_{ij}$) that allows stable coexistence of the three species without disturbance. In this simulation, we assumed $h_{ij} = h$ to reduce the dimension of parameter space. This analysis revealed that the stable coexistence of the three species is possible with a certain combinations of predation parameters. Within this parameter space, we used three parameter combinations to represent distinctive food webs with varied levels of omnivory: "chain (no omnivory)" [$(a_{BC}, a_{BP}, a_{CP}, h) = (0.5, 0, 0.0025, 0.5)$], "weak omnivory" ($0.5, 0.02, 0.025, 0.5$), and "strong omnivory" ($0.5, 0.04, 0.0025, 0.75$). We used $s = 1$ except for the chain scenario with $s = 0$. Without the loss of generality, we assumed $e_{ij} = e = 1$.

Under each scenario of omnivory, we crossed disturbance (frequency and intensity), productivity, and dispersal. Disturbance frequency and intensity were controlled via $p_m$ ( `r unique(df_sim$p_disturb) %>% sort()`) and $\mu_m$ (`r unique(df_sim$mean_disturb_source) %>% sort()`), respectively. Spatial variation in disturbance intensity at headwaters ($\sigma_{h}=$ `r max(df_sim$sd_disturb_source)`), and local disturbance variation ($\sigma_{l}=$ `r min(df_sim$sd_disturb_lon)`). This setup ($\sigma_h > \sigma_l$) produces high spatial variation among tributaries while allowing strong within-tributary downstream correlation, a typical pattern in real river networks [@nakamuraDisturbanceRegimesStream2000; @bendaNetworkDynamicsHypothesis2004]. We used the basal species' growth rate $r_b$ (`r unique(df_sim$r_b)`) as a proxy for resource availability as it resembles the gross primary production. For dispersal, we examined short- and long-distance dispersal scenarios (short-distance dispersal $\theta=$ `r max(df_sim$theta)`; long-distance dispersal $\theta=$ `r min(df_sim$theta)`) with constant dispersal probability ($p_{d}=$ `r unique(df_sim$p_dispersal)`).

We ran `r max(df_sim$n_rep)` simulation replicates with `r n_sim` timesteps of trophic dynamics (including `r n_sim - unique(df_sim$n_timestep)` steps of warmup and burn-in). During the warmup period (`r unique(df_sim$n_warmup)` time steps), we introduced propagules into habitat patches every 10 time steps to allow species to establish populations. Propagules are random draws from $\text{Poisson}(\lambda)$ ($\lambda=100, 10,~\text{and}~1$ for $B$, $C$, and $P$, respectively). Subsequently, we ran a burn-in period with no propagules for `r unique(df_sim$n_burnin)` time steps to remove the effects of initial conditions. We obtained the global FCL $\bar{\tau}$ for the last `r unique(df_sim$n_timestep)` time steps. Note that simulation replicates are differ in ecosystem size ($N_{p}\sim \text{Unif}(10,150)$) and ecosystem complexity ($P_{b} \sim \text{Unif}(0.01, 0.99)$) so we can estimate their effects on FCL. All analysis was performed in R 4.1.0 [@rcoreteamLanguageEnvironmentStatistical2021]
